Fast Sorting on Modern Computers

### 2005-08-26 docs/tech_report/background.tex
added bit about mispredictions vs misses, and about the scale of software prediction graphs
fixed from davids notes

@@

In the sequence T-T-T-N-T-T-T-N, a 2-bit predictor will only mispredict once.
In general, it will achieve 78 to 89\% accuracy.

@@

\section{Big O Notation}
Big O notation is used to describe the complexity of an algorithm. It says that
the algorithm will complete in a certain number of steps, in relation to number
of items of data being worked on, without giving specifics of how many
instructions each step will take to execute. It is possible, therefore, that an
algorithm with a lower complexity will complete slower than an algorithm where
each step is executed very quickly, but which has a greater order of complexity.

>>>>

Branch mispredictions are sometimes called \n{branch misses}, due to the
similarity between dynamic predictors and caches. In this report, we use the
terms interchangably.

@@

In the sequence T-T-T-N-T-T-T-N, a 2-bit predictor will only mispredict once per
loop iteration. In general, it will achieve 78 to 89\% accuracy.

@@

\section{Big O Notation}
Big O notation is used to describe the complexity of an algorithm. It says that
the algorithm will complete in a certain number of steps, in relation to number
of items of data being worked on, without giving specifics of how many
instructions each step will take to execute. It is possible, therefore, that an
algorithm with a lower complexity will complete more slowly than an algorithm
where each step is executed very quickly, but which has a greater order of
complexity.

### 2005-08-26 docs/tech_report/elementary.tex
fixed from davids notes

The level 1 cache should have bad performance as a result. Since it is smaller
than the array, the performance of the level 1 cache should simulate the
performance of the level 2 cache.

@@

Within practical sorting ranges, $H_N$ grows at a rate similar to
$log(N)$. Thus, in the average case, we can expect that the comparison
branch will mostly go in the same direction, making the branch highly
predictable.

@@

\code{Bubblesort code}{bubblesort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
These instruction count observations are again due to Sedgewick.  Bubblesort has
$N^2/2$ comparisons and exchanges. The number of passes over the array, and size
of each pass is the same as selection sort, and similar instruction count
results are expected.  The extra check to see if the array is sorted adds
$O(NlogN)$ instructions, but this is small in comparison to the number of
instructions saved if the sort ended early.

>>>>

The level 1 cache should have bad performance as a result. Since the level 1
cache is smaller that the arrays tested, it should perform analagiously to a
level 2 cache being tested with a much larger array.

@@

Within practical sorting ranges, $H_N$ grows at a rate similar to
$log(N)$\footnote{This seems like an error, since $H_N$ is very similar to
$log_eN$. It is not. We mean that $H_N$ is very similar to $log_2N$.}. Thus, in
the average case, we can expect that the comparison branch will mostly go in the
same direction, making the branch highly predictable.

@@

\code{Bubblesort code}{bubblesort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
These instruction count observations are again due to Sedgewick. Bubblesort has
$N^2/2$ comparisons and exchanges. The number of passes over the array, and size
of each pass is the same as selection sort, and similar instruction count
results are expected. The extra check to see if the array is sorted adds
$O(NlogN)$ instructions, but this is small in comparison to the number of
instructions saved if the sort ends early.

### 2005-08-26 docs/tech_report/method.tex
removed the word project, and resolved the occasional TODO
fixed from davids notes

Several steps were necessary to prepare for the research.  A framework was
required by which sorts could be tested as they were being developed. This is
discussed in Section \ref{Main}. It was also necessary to fully understand the
technologies to be used in the project. Several texts were used for this: cache
and branch predictors are both in \cite{HennessyPatterson90}; branch predictors
in \cite{Uht97}, as well as in \cite{McFarling93} and \cite{Smith81}; sorts are
discussed in \cite{Sedgewick02} and \cite{Knuth98}; cache-conscious sorts are
discussed \cite{LaMarca96}, \cite{LaMarca99} and \cite{LaMarcaHeap96}.

@@

\subsection{SimpleScalar} \n{SimpleScalar} is an architecture simulator. It
works by compiling code into its native instruction set, PISA, then emulating a
processor as it runs the program.  SimpleScalar can be found at
\n{http://www.simplescalar.com}. Documentation on how to use and modify
SimpleScalar are \cite{Burger97}, \cite{Austin02}, \cite{Tutorialv4},
\cite{Tutorialv2} and \cite{SimpleScalarUserGuide}.

SimpleScalar works by compiling code into its own instruction set. It provides a
compiler - a port of gcc - with related tools such as a linker and assembler.
Several simulators are then provided: \cc{sim-fast} only counts instruction,
quickly; \cc{sim-cache} simulates a cache and counts level 1 and level 2 cache
misses; \cc{sim-bpred} simulates a branch predictor and counts branch hits and
misses. Other simulators are provided, but they were not used in our tests.

@@

\subsection{PapiEx}
\cc{PapiEx} is a tool used to access performance counters on a Pentium 4. It
uses the PAPI framework to count certain characteristics across a program's
execution. We used it to count the number of cycles a sort took to run. This
provided an accurate measure of how fast a sort executes on a Pentium 4. The
PapiEx homepage is at \n{http://icl.cs.utk.edu/\textasciitilde{}mucci/papiex/}.

>>>>

Several steps were necessary to prepare for the research. A framework was
required by which sorts could be tested as they were being developed. This is
discussed in Section \ref{Main}. It was also necessary to fully understand the
technologies to be used in the report. Several texts were used for this: cache
and branch predictors are both in \cite{HennessyPatterson90}; branch predictors
in \cite{Uht97}, as well as in \cite{McFarling93} and \cite{Smith81}; sorts are
discussed in \cite{Sedgewick02} and \cite{Knuth98}; cache-conscious sorts are
discussed \cite{LaMarca96}, \cite{LaMarca99} and \cite{LaMarcaHeap96}.

@@

\subsection{SimpleScalar} \n{SimpleScalar} is an architecture simulator. It
works by compiling code into its native instruction set, \n{PISA}, then
emulating a processor as it runs the program. SimpleScalar can be found at
\n{http://www.simplescalar.com}. Documentation on how to use and modify
SimpleScalar are \cite{Burger97}, \cite{Austin02}, \cite{Tutorialv4},
\cite{Tutorialv2} and \cite{SimpleScalarUserGuide}.

SimpleScalar works by compiling code into its own instruction set. It provides a
compiler - a port of gcc - with related tools such as a linker and assembler.
Several simulators are then provided: \cc{sim-fast} only counts instructions
executed; \cc{sim-cache} simulates a cache and counts level 1 and level 2 cache
misses; \cc{sim-bpred} simulates a branch predictor and counts branch hits and
misses. Other simulators are provided, but they were not used in our tests.

@@

\subsection{PapiEx}
\n{PapiEx} is a tool used to access performance counters on a Pentium 4. It uses
the \n{PAPI} framework to count certain characteristics across a program's
execution. We used it to count the number of cycles a sort took to run. This
provided an accurate measure of how fast a sort executes on a Pentium 4. The
PapiEx homepage is at \n{http://icl.cs.utk.edu/\textasciitilde{}mucci/papiex/}.

### 2005-08-26 docs/tech_report/report.tex
removed the word project, and resolved the occasional TODO
added the numbers to the counter graphs
fixed from davids notes

% plots 6 or less images on a page, using the counters
\newcommand{\sixplot}[7]{
\afterpage{
\setcounter{subfigure}{0}
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-5em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[]{\label{#7-1}\includegraphics[scale=0.74]{plots/counter_#1}}
\ifthenelse{\equal{#2}{}}{}{\subfigure[]{\label{#7-2}\includegraphics[scale=0.74]{plots/counter_#2}}}
\ifthenelse{\equal{#3}{}}{}{\subfigure[]{\label{#7-3}\includegraphics[scale=0.74]{plots/counter_#3}}}
\ifthenelse{\equal{#4}{}}{}{\subfigure[]{\label{#7-4}\includegraphics[scale=0.74]{plots/counter_#4}}}
\ifthenelse{\equal{#5}{}}{}{\subfigure[]{\label{#7-5}\includegraphics[scale=0.74]{plots/counter_#5}}}
\ifthenelse{\equal{#6}{}}{}{\subfigure[]{\label{#7-6}\includegraphics[scale=0.74]{plots/counter_#6}}}
\end{changemargin}
\vspace*{6em}
\caption[Branch prediction performance for #7]{Branch prediction performance for #7. All graphs use the same scale.}
\label{Branch prediction performance for #7}
\end{figure}
\newpage
}}

@@

%TODO make the page wider
%TODO aspell in UK english
%TODO sometimes I say 3 sorts, sometimes 4. It's now 5.
%TODO do a final check for use of the word I
%TODO final check for the word 'project'

@@

This report describes a study of the behaviour of sorting algorithms
on branch predictors and caches. Our work on branch prediction is
almost entirely new, and finds a number of important results. In
particular we show that insertion sort causes the fewest branch
mispredictions of any comparison-based algorithm, that optimizations,
such as the choice of the pivot in quicksort, can have a large impact
on the predictability of branches, and that advanced two-level branch
predictors are usually worse at predicting branches in sorting
algorithms than simpler branch predictors. In many cases it is
possible to draw links between classical theoretical analyses of
algorithms and their branch prediction behaviour.

>>>>

% plots 6 or less images on a page, using the counters
\newcommand{\sixplot}[7]{
\afterpage{
\setcounter{subfigure}{0}
\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-5em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[]{\label{#7-1}\includegraphics[scale=0.74]{plots/counter_#1}}
\ifthenelse{\equal{#2}{}}{}{\subfigure[]{\label{#7-2}\includegraphics[scale=0.74]{plots/counter_#2}}}
\ifthenelse{\equal{#3}{}}{}{\subfigure[]{\label{#7-3}\includegraphics[scale=0.74]{plots/counter_#3}}}
\ifthenelse{\equal{#4}{}}{}{\subfigure[]{\label{#7-4}\includegraphics[scale=0.74]{plots/counter_#4}}}
\ifthenelse{\equal{#5}{}}{}{\subfigure[]{\label{#7-5}\includegraphics[scale=0.74]{plots/counter_#5}}}
\ifthenelse{\equal{#6}{}}{}{\subfigure[]{\label{#7-6}\includegraphics[scale=0.74]{plots/counter_#6}}}
\end{changemargin}
\vspace*{6em}
\caption[Branch prediction performance for #7]{Branch prediction performance for
#7. All graphs use the same scale. Above each column is the total number of
branches and the percentage of correctly predicted branches for a simulated
bimodal predictor.}
\label{Branch prediction performance for #7}
\end{figure}
\newpage
}}

@@

%TODO make the page wider

@@

This report describes a study of the behaviour of sorting algorithms
on branch predictors and caches. Our work on branch prediction is
almost entirely new, and finds a number of important results. In
particular we show that insertion sort causes the fewest branch
mispredictions of any comparison-based algorithm, that optimizations -
such as the choice of the pivot in quicksort - can have a large impact
on the predictability of branches, and that advanced two-level branch
predictors are usually worse at predicting branches in sorting
algorithms than simpler branch predictors. In many cases it is
possible to draw links between classical theoretical analyses of
algorithms and their branch prediction behaviour.
